---
title: "Pseudobulk differential expression with DESeq2"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 5
vignette: >
    %\VignetteIndexEntry{Pseudobulk differential expression with DESeq2}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
)
options(digits = 3)
```

# Why pseudobulk?

Single cells from the same donor are not statistically independent: they
share genetic background, batch effects, and ambient RNA. Treating each
cell as an independent observation in a regression inflates the effective
sample size and distorts p-values.

A common, principled remedy is **pseudobulk**: sum each donor's cells
within a group (e.g. cluster) to produce a per-donor count vector, then
run a count-based regression like `DESeq2::DESeq` over donors. The unit
of replication becomes the donor, not the cell.

`presto` provides:

- `collapse_counts()` — fast collapse of a cell-by-gene count matrix to
  pseudobulks defined by one or more metadata columns.
- `pseudobulk_deseq2()` — run DESeq2 across the resulting pseudobulks in
  a one-vs-all, pairwise, or within-group design.
- `top_markers_dds()` — extract top features per group from the result.

# Demo data

We use the colon-tissue CD8 T-cell dataset from
[Thomas et al. *Nat. Med.* 2024](https://www.nature.com/articles/s41591-024-02895-x)
([GSE206299](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE206299)).
27 donors, 9 cell clusters, 25,341 cells.

```{r load}
library(presto)
library(Matrix)
library(dplyr)

d <- load_ircolitis_cd8(verbose = FALSE)
dim(d$counts)
table(d$obs$donor, d$obs$cluster)[1:5, ]
```

The crosstab confirms most donors contribute cells to most clusters —
good news for pseudobulk, since clusters with few represented donors
will have low statistical power.

# Step 1: collapse cells to pseudobulks

`collapse_counts()` sums the columns of a count matrix according to one
or more metadata columns. Here we collapse by `(donor, cluster)`, drop
any (donor, cluster) pair with fewer than 10 cells, and keep the cell
count `N` so we can inspect what came through.

```{r collapse}
md <- data.frame(
    donor   = as.character(d$obs$donor),
    cluster = as.character(d$obs$cluster),
    case    = as.character(d$obs$case),
    stringsAsFactors = FALSE
)
cc <- collapse_counts(
    d$counts, md, c("donor", "cluster", "case"),
    min_cells_per_group = 10,
    keep_n = TRUE
)
dim(cc$counts_mat)
head(cc$meta_data)
```

We collapse by `(donor, cluster, case)` instead of just `(donor, cluster)`
because `case` is constant within donor — including it now makes it
available for the within-cluster contrast in
[Step 4](#step-4-within-cluster-case-vs-control).

`cc$counts_mat` is now a 28,165-gene × `r ncol(cc$counts_mat)`-pseudobulk
matrix. Each column is one (donor, cluster) sample; `cc$meta_data` is the
matching `data.frame` of pseudobulk identities.

```{r per-cluster}
table(cc$meta_data$cluster)
```

# Step 2: one-vs-all DESeq2

`pseudobulk_deseq2()` with `mode = "one_vs_all"` tests each cluster
against the union of the others. With `collapse_background = FALSE` we
keep each background pseudobulk as its own observation; setting it to
`TRUE` collapses the background pseudobulks per donor first, which is
preferable when donors are unbalanced across clusters.

The first variable in the formula is the contrast variable. `meta_data`
should contain only the pseudobulk identifying variables used in the
formula.

```{r dge}
md_dge <- cc$meta_data[, "cluster", drop = FALSE]
res <- pseudobulk_deseq2(
    ~cluster, md_dge, cc$counts_mat,
    verbose = FALSE,
    min_counts_per_sample = 10,
    present_in_min_samples = 5,
    collapse_background = FALSE,
    mode = "one_vs_all"
)
head(res)
```

Result columns come straight from `DESeq2::results()`, prepended with the
group identifier:

| column          | description                                       |
|-----------------|---------------------------------------------------|
| `group`         | foreground cluster                                 |
| `feature`       | gene identifier                                    |
| `baseMean`      | mean of normalized counts                          |
| `log2FoldChange`| effect size, group vs background                   |
| `lfcSE`         | standard error of `log2FoldChange`                 |
| `stat`          | Wald statistic                                     |
| `pvalue`        | nominal p-value                                    |
| `padj`          | Benjamini-Hochberg adjusted p-value                |

# Step 3: top markers per cluster

`top_markers_dds()` filters by `padj_max` and `lfc_min` and returns a
wide table of the top features per group:

```{r top}
top5 <- top_markers_dds(res, n = 5, padj_max = 1e-4, lfc_min = 1)
top5 %>%
    dplyr::mutate(across(-rank, ~ sub(".*\\|", "", .x)))
```

The signal is biologically reasonable: cluster 1 is dominated by
S-phase / DNA-replication genes (`DUT`, `LIG1`, `H2AFX`, `TMPO`),
cluster 2 by effector chemokines (`CCL4`, `CCL3L1`).

# Step 4: pairwise cluster contrasts

`mode = "pairwise"` tests each cluster against each other cluster
individually, returning a long-form table with `group1` and `group2`
columns. This is more conservative than one-vs-all because a true
marker has to differentiate the cluster from *every* other cluster, not
just the average background.

The cost grows quickly — *N* clusters means *N × (N−1)* directional
DESeq2 fits — so it's worth subsetting to the levels you care about
first. Here we contrast clusters 1, 2, and 3:

```{r pairwise}
keep <- cc$meta_data$cluster %in% c("1", "2", "3")
md_p  <- cc$meta_data[keep, "cluster", drop = FALSE]
mat_p <- cc$counts_mat[, keep]

res_p <- pseudobulk_deseq2(
    ~cluster, md_p, mat_p,
    verbose = FALSE,
    min_counts_per_sample = 10,
    present_in_min_samples = 5,
    mode = "pairwise"
)
head(res_p)
```

`summarize_dge_pairs()` collapses the directional pairs to one row per
(group, gene). Pass `"min"` to keep each gene's worst comparison (the
*most conservative* effect, useful for high-confidence markers) or
`"max"` for the best.

```{r summarize-pairs}
summarize_dge_pairs(res_p, "min") %>%
    head(10) %>%
    dplyr::mutate(feature = sub(".*\\|", "", feature))
```

# Step 5: within-cluster Case vs Control

A more interesting biological question than "what defines each cluster"
is "for each cluster, which genes change between irColitis cases and
healthy controls?". `mode = "within"` answers this. The first variable
in the formula is the **split variable** (the cluster), and the second
is the **contrast variable** (`case`). Within each level of the split
variable, DESeq2 fits the inner formula (here `~case`).

```{r within}
md_w <- cc$meta_data[, c("cluster", "case")]

res_w <- pseudobulk_deseq2(
    ~cluster + case, md_w, cc$counts_mat,
    verbose = FALSE,
    min_counts_per_sample = 10,
    present_in_min_samples = 5,
    mode = "within"
)
head(res_w)
```

Top genes upregulated in cases per cluster:

```{r within-top}
top_markers_dds(res_w, n = 5, padj_max = 1e-4, lfc_min = 1) %>%
    dplyr::mutate(across(-rank, ~ sub(".*\\|", "", .x)))
```

Cluster-1 cases up-regulate `NCF1`, `PRNP`, `TXNIP` — interferon-response
and oxidative-stress genes consistent with inflamed tissue.

# Tips

- **Filter low-count genes early.** `min_counts_per_sample` and
  `present_in_min_samples` are applied before fitting; tighter filters
  speed up the DESeq2 step substantially.
- **Always include batch / channel covariates when possible.** Add them
  to the formula after the contrast variable (e.g. `~cluster + channel`)
  if the design supports it.
- **`collapse_background = TRUE`** (in `mode = "one_vs_all"`) can
  stabilize results when donor representation is uneven across clusters.
- **`mode = "within"` supports both 2-level and ordinal contrasts.**
  Two-level contrasts (factor or character) get a Wald test on the
  level-vs-reference coefficient. Three or more levels are
  integer-encoded and treated as an ordinal trend.

# Session info

```{r}
sessionInfo()
```
